Simple Wind Models

### 2010-08-02 Winds/papers/Theory/First/wind_simple.tex
blah
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5095 ac11df5e-a788-c23b-a8e1-c758c323924d

\newcommand{\bturb}{$b_{\rm turb}$}
\newcommand{\mbturb}{b_{\rm turb}}
\newcommand{\taud}{$\tau_{\rm dust}$}
\newcommand{\mtaud}{\tau_{\rm dust}}
\newcommand{\maconfig}{2p$^6$3s}
\newcommand{\mbconfig}{2p$^6$3p}
\newcommand{\aconfig}{a~$^6$D$^0$}
\newcommand{\zconfig}{z~$^6$D$^0$}
\newcommand{\mvr}{v_{\rm r}}
\newcommand{\naid}{\ion{Na}{1}~$\lambda\lambda 5891, 5897$}
\newcommand{\mgiid}{\ion{Mg}{2}~$\lambda\lambda 2796, 2803$}
\newcommand{\mgiia}{\ion{Mg}{2}~$\lambda 2796$}
\newcommand{\mgiib}{\ion{Mg}{2}~$\lambda 2803$}
\newcommand{\feiia}{\ion{Fe}{2}~$\lambda 2586$}
\newcommand{\feiib}{\ion{Fe}{2}~$\lambda 2600$}
\newcommand{\feiic}{\ion{Fe}{2}$^* \; \lambda 2612$}
\newcommand{\feiie}{\ion{Fe}{2}$^* \; \lambda 2626$}
\newcommand{\feiid}{\ion{Fe}{2}~$\lambda\lambda 2586, 2600$}
\newcommand{\feiis}{\ion{Fe}{2}$^*$}
\newcommand{\nmg}{$n_{\rm Mg^+}$}
\newcommand{\mnmg}{n_{\rm Mg^+}}
\newcommand{\nfe}{$n_{\rm Fe^+}$}
\newcommand{\mnfe}{n_{\rm Fe^+}}
\def\hub{h_{72}^{-1}}
\def\umfp{{\hub \, \rm Mpc}}
\def\mzq{z_q}
\def\zabs{$z_{\rm abs}$}
\def\mzabs{z_{\rm abs}}
\def\intl{\int\limits}
\def\cmma{\;\;\; ,}
\def\perd{\;\;\; .}
\def\ltk{\left [ \,}
\def\ltp{\left ( \,}
\def\ltb{\left \{ \,}
\def\rtk{\, \right  ] }
\def\rtp{\, \right  ) }
\def\rtb{\, \right \} }
\def\sci#1{{\; \times \; 10^{#1}}}
\def \rAA {\rm \AA}
\def \zem {$z_{\rm em}$}
\def \mzem {z_{\rm em}}
\def\smm{\sum\limits}
\def \cmm  {cm$^{-2}$}
\def \cmmm {cm$^{-3}$}
\def \kms  {km~s$^{-1}$}
\def \mkms  {\, {\rm km~s^{-1}}}
\def \lyaf {Ly$\alpha$ forest}
\def \Lya  {Ly$\alpha$}
\def \lya  {Ly$\alpha$}
\def \mlya  {Ly\alpha}
\def \Lyb  {Ly$\beta$}
\def \lyb  {Ly$\beta$}
\def \lyg  {Ly$\gamma$}
\def \ly5  {Ly-5}
\def \ly6  {Ly-6}
\def \ly7  {Ly-7}
\def \nhi  {$N_{\rm HI}$}
\def \mnhi  {N_{\rm HI}}
\def \lnhi {$\log N_{HI}$}
\def \mlnhi {\log N_{HI}}
\def \etal {\textit{et al.}}
\def \lyaf {Lyman--$\alpha$ forest}
\def \mnmin {\mnhi^{\rm min}}
\def \nmin {$\mnhi^{\rm min}$}
\def \O {${\mathcal O}(N,X)$}
\newcommand{\cm}[1]{\, {\rm cm^{#1}}}
\def \snrlim {SNR$_{lim}$}

@@

\section{Observables}
\label{sec:obs}

Although the focus of this paper is to explore the absorption/emission
profiles for idealized wind models, we ultimately intend to compare 
profiles like these against observations to constrain characteristics
of the wind (e.g.\ radial extent, energetics).  In this section we
consider a few observables and their dependence on the wind properties.

\subsection{Absorption profiles}

 \begin{enumerate}
   \item Peak optical depth
     \begin{itemize}
       \item Not a robust measure of $\tau_{wind}$
       \item Not even a good relative predictor for $\tau$
       \item Not a robust predictor of $f_c$
       \item Scatter plot of $\tau$ inferred vs.\ $\tau$ true
     \end{itemize}
   \item Kinematics
     \begin{itemize}
       \item Nearly robust for $v<v_{min}$
       \item But what sets $v_{min}$?
       \item Generally get the full extent of the wind speed
       \item Plot $v_{pk}$ vs $\tau_{pk}$
     \end{itemize}
   \item EW
     \begin{itemize}
       \item Primarily set by $v_{min}(\tau \sim 1)$
       \item $f\lambda$ matters, but so do emission channels
       \item (EW measured - EW intrinsic) vs.\ EW intrinsic
     \end{itemize}
   \item Fine-structure
     \begin{itemize}
       \item Distance to the source
     \end{itemize}
 \end{enumerate}

\subsection{Emission profiles}

 \begin{enumerate}
   \item Resonant lines
     \begin{itemize}
       \item Kinematics of backside
       \item Flux + relative
       \item Plot $v_{pk}$ emission vs.\ $v_{pk}$ absorption?
    \end{itemize}
   \item Fine-structure
     \begin{itemize}
       \item Flux $\propto \tau_{wind}$ + relative
       \item Kinematics (anisotropy)
    \end{itemize}
 \end{enumerate}

\section{Discussion}
\label{sec:discuss}

@@

>>>>

\newcommand{\ewabs}{$W_{\rm a}$}
\newcommand{\ewe}{$W_{\rm e}$}
\newcommand{\rtt}{$r_{\tau = 0.2}$}
\newcommand{\mrtt}{r_{\tau = 0.2}}
\newcommand{\bturb}{$b_{\rm turb}$}
\newcommand{\mbturb}{b_{\rm turb}}
\newcommand{\taud}{$\tau_{\rm dust}$}
\newcommand{\mtaud}{\tau_{\rm dust}}
\newcommand{\maconfig}{2p$^6$3s}
\newcommand{\mbconfig}{2p$^6$3p}
\newcommand{\aconfig}{a~$^6$D$^0$}
\newcommand{\zconfig}{z~$^6$D$^0$}
\newcommand{\mvr}{v_{\rm r}}
\newcommand{\naid}{\ion{Na}{1}~$\lambda\lambda 5891, 5897$}
\newcommand{\mgiid}{\ion{Mg}{2}~$\lambda\lambda 2796, 2803$}
\newcommand{\mgiia}{\ion{Mg}{2}~$\lambda 2796$}
\newcommand{\mgiib}{\ion{Mg}{2}~$\lambda 2803$}
\newcommand{\feiia}{\ion{Fe}{2}~$\lambda 2586$}
\newcommand{\feiib}{\ion{Fe}{2}~$\lambda 2600$}
\newcommand{\feiic}{\ion{Fe}{2}$^* \; \lambda 2612$}
\newcommand{\feiie}{\ion{Fe}{2}$^* \; \lambda 2626$}
\newcommand{\feiid}{\ion{Fe}{2}~$\lambda\lambda 2586, 2600$}
\newcommand{\feiis}{\ion{Fe}{2}$^*$}
\newcommand{\nmg}{$n_{\rm Mg^+}$}
\newcommand{\mnmg}{n_{\rm Mg^+}}
\newcommand{\nfe}{$n_{\rm Fe^+}$}
\newcommand{\mnfe}{n_{\rm Fe^+}}
\def\hub{h_{72}^{-1}}
\def\umfp{{\hub \, \rm Mpc}}
\def\mzq{z_q}
\def\zabs{$z_{\rm abs}$}
\def\mzabs{z_{\rm abs}}
\def\intl{\int\limits}
\def\cmma{\;\;\; ,}
\def\perd{\;\;\; .}
\def\ltk{\left [ \,}
\def\ltp{\left ( \,}
\def\ltb{\left \{ \,}
\def\rtk{\, \right  ] }
\def\rtp{\, \right  ) }
\def\rtb{\, \right \} }
\def\sci#1{{\; \times \; 10^{#1}}}
\def \rAA {\rm \AA}
\def \zem {$z_{\rm em}$}
\def \mzem {z_{\rm em}}
\def\smm{\sum\limits}
\def \cmm  {cm$^{-2}$}
\def \cmmm {cm$^{-3}$}
\def \kms  {km~s$^{-1}$}
\def \mkms  {\, {\rm km~s^{-1}}}
\def \lyaf {Ly$\alpha$ forest}
\def \Lya  {Ly$\alpha$}
\def \lya  {Ly$\alpha$}
\def \mlya  {Ly\alpha}
\def \Lyb  {Ly$\beta$}
\def \lyb  {Ly$\beta$}
\def \lyg  {Ly$\gamma$}
\def \ly5  {Ly-5}
\def \ly6  {Ly-6}
\def \ly7  {Ly-7}
\def \nhi  {$N_{\rm HI}$}
\def \mnhi  {N_{\rm HI}}
\def \lnhi {$\log N_{HI}$}
\def \mlnhi {\log N_{HI}}
\def \etal {\textit{et al.}}
\def \lyaf {Lyman--$\alpha$ forest}
\def \mnmin {\mnhi^{\rm min}}
\def \nmin {$\mnhi^{\rm min}$}
\def \O {${\mathcal O}(N,X)$}
\newcommand{\cm}[1]{\, {\rm cm^{#1}}}
\def \snrlim {SNR$_{lim}$}

@@

%\section{Observables}
%\label{sec:obs}
%
%%Although the focus of this paper is to explore the absorption/emission
%profiles for idealized wind models, we ultimately intend to compare 
%profiles like these against observations to constrain characteristics
%of the wind (e.g.\ radial extent, energetics).  In this section we
%consider a few observables and their dependence on the wind properties.
%
%\subsection{Absorption profiles}
%
% \begin{enumerate}
%   \item Peak optical depth
%     \begin{itemize}
%       \item Not a robust measure of $\tau_{wind}$
%       \item Not even a good relative predictor for $\tau$
%       \item Not a robust predictor of $f_c$
%       \item Scatter plot of $\tau$ inferred vs.\ $\tau$ true
%     \end{itemize}
%   \item Kinematics
%     \begin{itemize}
%       \item Nearly robust for $v<v_{min}$
%       \item But what sets $v_{min}$?
%       \item Generally get the full extent of the wind speed
%       \item Plot $v_{pk}$ vs $\tau_{pk}$
%     \end{itemize}
%   \item EW
%     \begin{itemize}
%       \item Primarily set by $v_{min}(\tau \sim 1)$
%       \item $f\lambda$ matters, but so do emission channels
%       \item (EW measured - EW intrinsic) vs.\ EW intrinsic
%     \end{itemize}
%   \item Fine-structure
%     \begin{itemize}
%       \item Distance to the source
%     \end{itemize}
% \end{enumerate}
%     
%\subsection{Emission profiles}
%
% \begin{enumerate}
%   \item Resonant lines
%     \begin{itemize}
%       \item Kinematics of backside
%       \item Flux + relative
%       \item Plot $v_{pk}$ emission vs.\ $v_{pk}$ absorption?
%    \end{itemize}
%   \item Fine-structure
%     \begin{itemize}
%       \item Flux $\propto \tau_{wind}$ + relative
%       \item Kinematics (anisotropy)
%    \end{itemize}
% \end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:discuss}

The previous sections presented a series of idealized wind models for
cool gas outflows, and explored the absorption/emission profiles of
the \mgiid\ doublet and the \ion{Fe}{2} UV1 multiplet.  In addition to
the ubiquitous presence of blue-shifted absorption, the wind models
also generically predict strong line emission in both the resonance
and non-resonance transitions.  For isotropic and dust-free scenarios,
this is simple convservation of photon number: every photon emitted by
the source eventually escapes the system with the wind only [slightly]
shifting them in frequency space.  A principal result of this paper,
therefore, is that cool gas outflows should generate detectable line
emission.

As noted in the introduction, there are several reports of line
emission associated with star-forming galaxies that exhibit cool gas outflows.  
[SHORTEN THE OBSERVER SUMMARY]
[Phillips first reported?]
Bright \ion{Mg}{2} and \ion{Fe}{2}* line emission has also been
reported for individual galaxies at $z \sim 0.7$ by \cite{rubin+10b}
and \cite{rubin+10c}.   
\cite{wcp+09}, who
studied \ion{Mg}{2} absorption in $z \sim 1.4$ galaxies, reported
\ion{Mg}{2} emission in a small subset of the individual galaxy
spectra of their large sample.  These were excluded from the full
analysis on concerns that the emission was related to AGN activity.
The stacked spectra of the remaining galaxies, however, also indicated
\ion{Mg}{2} emisison, both directly and when the authors modeled and
`removed' the $v \approx 0\kms$ absorption component.  Although the
authors suggested the emission could be related to back-scattered
light in the wind, they also allowed that it could be related to weak
AGN activity.   Very similar \ion{Mg}{2} emission was observed by
\cite{rubin+10b} who repeated the analysis of \cite{wcp+09} on a set
of lower redshift galaxies. 
[Did Kate see FeII* emission in her stacks? If not, why not?]
[Are there aspects of those works that are now understood, e.g. the
difference in FeII 2586 and FeII 2600]
At high redshift, \ion{Si}{2}* line emission has been reported from
the Lyman break galaxies which was suggested to be
related to \ion{H}{2} regions within the galaxies \citep{shapley03}.
Lastly, \ion{Na}{1} line emission has also been detected in the
stacked spectra of $z \sim 0$ galaxies by \cite{chen}.  
We contend that all of these incidences may be examples of line
scattered photons in the cool gas outflows of these star-forming
galaxies.  Indeed, our results indicate that line emission could (and
should) be produced by all of these galaxies if the outflows (and the
galaxy emission) are nearly isotropic.

On the other hand,
there are many examples of galaxies where blue-shifted absorption is
detected yet the authors report no significant line emission.  This
includes the \ion{Na}{1} flows reported by \cite{rupke} and
\cite{martin}, the \ion{Mg}{2} and \ion{Fe}{2} transitions studied in
ULIRGS by \cite{mb10}, and the extreme \ion{Mg}{2} outflows identified
by \cite{tremonti}.  Similarly, there have been no reports of
resonance line emission from the LBGs nor ... [Others]
The multitude of non-detections stands in stark contradiction to the
results presented in this paper.   We are motivated, therefore, to
reassess several of the effects that may reduce the line emission
and consider whether these may eliminate the line emission for many
star-forming galaxies.  [If MgII emission is 'common' for HII regions,
why don't we see it *always* when we see OII, Hbeta, etc?  Same for
FeII*]

Dust is frequently invoked to explain the suppression of line emission
related to a resonantly trapped line (e.g.\ \lya).  Indeed, a photon
that is trapped for many scatterings within a dust medium will be
preferentially extincted relative to a non-resonant photon.  In
$\S$~\ref{sec:dust}, we examined the effects of dust extinction on the
\ion{Mg}{2} and \ion{Fe}{2} emission.  The general result was a modest
suppression of the line flux of the order of (1+\taud)$^{-1}$ instead of
exp(--\taud).   [Do a series of models and fit a functional form for
the fiducial model]  Although the \ion{Mg}{2} photons are resonantly
trapped within the outflow, they require only one to a few scatterings
to escape the fiducial wind model and dust has only a
modest effect.  This reflects the [modest] opacity of the \ion{Mg}{2}
doublet (e.g.\ relative to \lya) but also the specific characteristics
of the wind.  We have also experimented with dust in several of the
other models (e.g.\ the ISM+wind scenario) and find similar results.
[Dust does reduce the $v \sim 0$ emission in the ISM+wind model more,
as it is more effectively trapped.  Discuss this somewhere..  Also
note that the EW of the absorption is higher for an ISM model so that
the absolute value of EW emission is higher too for FeII*]
Another reason \ion{Mg}{2} is less susceptible than \lya\ is that the
latter may be generated through the flourescence of an ionizing
photon, i.e.\ via a recombination cascade.  The ionizing photons 
would penetrate deeper into the outflow and more effectively trap the
\lya\ photon [is this right?].

We also emphasize that the \ion{Fe}{2} resonance photons, especially
\feiia, are converted to non-resonant photons and escape the wind
freely.  A wind model that effectively trapped \ion{Mg}{2}
photons for many scatterings would not be expected to trap the
\feiid\ photons.  Therefore, the effects of dust are 
reduced for \ion{Fe}{2}* and, by inference, any other set of
transitions that are coupled to a fine-structure level (e.g.\
\ion{Si}{2} and \ion{Si}{2}*).  
Although dust does reduce the line flux relative to the continuum, we
conclude that it has only a modest effect on \ion{Mg}{2} and
\ion{Fe}{2}* emission.

Another factor that reduces the predicted line emission is a
non-isotropic wind model,e.g.\ the flux is reduced if there is no
backside of the wind.  To zeroth order, the line emission scales
roughly as $\Omega/4\pi$ where $\Omega$ is the angular extent of the
wind\footnote{There is also a first order dependence on the actual
  orientation of the wind with emission maximal if the backside is
  fully present.}.  Because we require that the wind is at least
pointing towards us, it is difficult to reduce $\Omega$ signficantly
below $2 \pi$.  Therefore, we expect this effect to reduce the
emission by a factor of order unity.  
[The notion of the source absorbing all photons that scatter into it
is akin to removing the backside]

By a similar [token], an anisotropic source could alter the predicted
line emission.  For example, if the backside of the galaxy were
brighter/fainter than would one predict brighter/fainter emission
relative to the observed continuum.
To reduce the emission, the brightest regions
of the galaxy would need to be oriented toward Earth.  Although this would not a
generic occurance, most spectroscopic samples are magnitude-limited
and therefore biased to detecting galaxies at their brightest.

Together, dust and non-isotropic wind models can reduce line emission
by a factor of a few.  We suggest, however, that this would generally
not be sufficient to reduce the emission below the typical detection
limits.

There is another, more subtle effect that in principle will
substantially reduce the observed line emission: slit losses.  As
described in Figures~\ref{fig:fiducial_ifu_mgii} and
\ref{fig:ISM_ifu}, emission from the wind is spatially extended with
at least modest surface brightness at large radii implying a
non-negligible total luminosity from well beyond the galaxy.  
The majority of observations of star-forming galaxies to date have
been taken through a spectroscopic slit typically designed to cover
the brightest continuum regions of the galaxy.  A slit with an angular
extent of $1-1.5''$ subtends roughly $8-15$\,kpc for galaxies at $z>0.5$.
Therefore, a slit covering a galaxy with our fiducial wind would cover
less than half of the wind.  The result will be reduced line emission,
especially in comparison with the galaxy continuum.

We have investigated the effect of a [partial] aperture as follows.
We model the slit as a perfect, infinte rectangle centered on the
galaxy.  We then tabulate the equivalent width of the line emission
through slits with a range of widths, parameterized by the radius
where the wind has an effective optical depth in the line of $\tau =
0.2$ (\rtt)\footnote{Specifically, we determine the
  velocity where $\tau(v) = 0.2$ and then use the velocity law to
  convert to a radius.}.   We expect the line emission is weak beyond
this radius because the photons have only a low probability of
interacting with the wind.  For the fiducial wind model, $\mrtt
\approx 15$\,kpc for \mgiia\ (Figure~\ref{fig:fiducial_nvt}).

The predicted equivalent width in emission \ewe\ relative to the absorption
equivalent width \ewabs\ is presented in
Figure~\ref{fig:obs_slit} for a set of transitions (black curves).
The curves rise very steeply with increasing slit width and then
plateau when the slit width is $\approx \mrtt$.  
The results are comparable for other wind models considered
in this paper (colored curves in Figure~\ref{fig:obs_slit}).
We conclude that slit losses would be important for slits that do not
cover a majority of the wind.   Again, the effect is of order unity
unless the slit width is very small.   Provided the winds do not
extend far beyond 10\,kpc, the majority of emission will be subtended
by stnadard slit observations.  Nevertheless, the results do motivate
the usage of extended apertures, e.g.\ integral field unit (IFU)
instrumentation that would map the wind both spatially and spectrally.

Although there are several effects that will reduce line emission
relative to the absorption produced by the outflow, our analysis
indicates that detectable line emission should be present for [the
majority] of winds.  One of the principal effects of this emission is
that it modifies the observed absorption profiles (e.g.\
Figure~\ref{fig:noemiss}) which may lead to erroneous conclusions
regarding characteristics of the flow.  {\it Analysis of cool gas outflows
that ignores emission may incorrectly lead to conclusions the gas is
partially covered, has a significantly lower peak optical depth,
and/or that a $v \sim 0 \mkms$ ISM component (perhaps even infall) is
absent.}  Figure~\ref{fig:obs_ew} demonstrates a simple observational
consequence of the line emission: reduced estimates for the absorption
equivalent width \ewabs\ of the flow.  In the case of \ion{Mg}{2},
which has the most affected transitions, the \ewabs\ is reduced by
XX\% for the wind models presented in this paper.

[Could they have accidentally `sky subtracted' it?  I doubt it]

%%%%%%%%%%%%%%%%%

@@

\begin{figure}
\includegraphics[scale=0.6,angle=90]{Figures/fig_obs_slit.ps}
\caption{
Slit width effects
}
\label{fig:obs_slit}
\end{figure}

\begin{figure}
\includegraphics[scale=0.6,angle=90]{Figures/fig_obs_ew.ps}
\caption{
EW effect
}
\label{fig:obs_ew}
\end{figure}
